---
title: "Overview"
author: "Randy Johnson"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reshape2)
library(tidyverse)
library(lubridate)
```

# Introduction

This document gives an overview that broadly covers the following information:

* What is tidy data?
* Why use the `tidyverse`?
    * `%>%`
    * `mutate`
    * `tibble`
    * miscelaneous data processing tools in the `tidyverse`

# What is tidy data?

In short, tidy data has one row per observation, and one column per measurement. For example, whe `who` data set is well organized and compact, but it is does not meet the definition of tidy:

```{r not tidy who}
# load the who data set from the tidyr package
require(tidyr)
data(who)

# display who
who
```

The `who` data set has information from thw World Health Organization on TB cases recorded by country, age and sex. It is not tidy because multiple variables are encoded in many of the columns, and incidence is spread over columns 5 through 60 (e.g. `new_sp_m014` contains incidence for males between 0 and 14 years of age, who were diagnosed using a positive pulmonary smear). To covert this to a tidy format we could do the following:

```{r tidy who}
# save country codes in a separate table
isoCodes <- select(who, country, iso2, iso3) %>%
            unique()

# tidy the who data set
        # gather incidence into one column
who2 <- gather(who, 'group', 'incidence', -country, -iso2, -iso3, -year) %>%
 
        # spread group out into separate columns
        mutate(group = str_replace(group, 'new_?', ''),        # remove "new" and "new_" from the group name
                        
               diagnosis = str_split_fixed(group, '_', 2)[,1], # grab the diagnosis type
               
               sexAgeCat = str_split_fixed(group, '_', 2)[,2], # grab sex/age category
               sex = substr(sexAgeCat, 1, 1),                  #    first character is sex
               ageCat = substr(sexAgeCat, 2, 5)) %>%           #    next 4 characters are age category
    
        # drop rows with missing incidence
        filter(!is.na(incidence)) %>%

        # drop redundant columns
        select(-group, -iso2, -iso3, -sexAgeCat)

who2
```

For some additional information on tidy data, see [Ravi's slides](blob/master/TidyingData-StatForLunch.pdf) from his Statistics for Lunch talk.

# Why use the `tidyvese`?

There are many reasons to use `tidyverse` - I've listed my favorite parts of the `tidyverse` here, but there are many more that we don't have room to discuss. For starters, `library(tidyverse)` will load the following packages (it is really an empty package with a bunch of dependencies):

* ggplot2: This is a completely new way of generating graphics in R that is based on the grammar of graphics.
* tibble: This implements a reimmagining of the `data.frame`. The `tibble` or `data_frame` is fairly backward compatible with functions that work with the `data.frame`. The only exception I know of is the `missForest` package.
* tidyr: Provides `gather` and `spread` functions for data tidying.
* readr: Redesigned import functions for reading csv, tsv... text data files.
* purrr: New tools, similar to the `apply` family, for working with the `tibble`.
* dplyr: Data manipulation tools.
* stringr: String manipulation tools.
* forcats: Categorical data (i.e. `factor`) manipulation tools.
* magrittr: This packge isn't actually loaded, but the pipe operator, `%>%`, is imported.

There are a number of additional dependencies that are installed when you use the `dependencies = TRUE` option (click on the links for more details):

* [broom](https://github.com/tidyverse/broom)
* [cli](https://github.com/r-lib/cli)
* [crayon](https://github.com/r-lib/crayon)
* [dbplyr](https://github.com/tidyverse/dbplyr)
* [haven](https://github.com/tidyverse/haven)
* [hms](https://github.com/tidyverse/hms)
* [httr](https://cran.r-project.org/web/packages/httr/index.html)
* [jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html)
* [lubridate](https://github.com/tidyverse/lubridate)
* [modelr](https://github.com/tidyverse/modelr)
* [readxl](https://github.com/tidyverse/readxl)
* [reprex](https://github.com/tidyverse/reprex)
* [rlang](https://cran.r-project.org/web/packages/rlang/index.html)
* [rstudioapi](https://cran.rstudio.com/web/packages/rstudioapi/index.html)
* [rvest](https://cran.r-project.org/web/packages/rvest/rvest.pdf)
* [xml2](https://cran.r-project.org/web/packages/xml2/index.html)
* [feather](https://blog.rstudio.com/2016/03/29/feather/)
* [knir](https://cran.r-project.org/web/packages/knitr/index.html)
* [rmarkdown](https://cran.r-project.org/web/packages/markdown/index.html)

## Pipes

The `magrittr` package was merged into the `tidyverse` early in its development, and the `tidyverse` makes extensive use of the pipe operator, `%>%`. Compare the code above to this code used to tidy the `who` data set.

```{r tidy who2}
# drop redundant country information
who3 <- subset(who, select = c(1, 4:60))

##### tidy the who data set #####
# melt who3
require(reshape2)
who3 <- melt(who3, id = c('country', 'year'), na.rm = TRUE, value.name = 'incidence')

# split `variable` into diagnosis/sexAgeCat vectors
tmp <- strsplit(gsub("new_?", "", as.character(who3$variable)), "_")

who3$diagnosis <- sapply(tmp, `[`, 1)            # get diagnosis
who3$sex <- substr(sapply(tmp, `[`, 2), 1, 1)    # get sex
who3$ageCat <- substr(sapply(tmp, `[`, 2), 2, 5) # get age category

# get rid of `variable`
who3$variable <- NULL

str(who3)
```

* `tidyverse` functions were built with `%>%` in mind, so we always know that the first argument is data. This makes passing output from one function to the next easy. Other traditional R functions are inconsistent (e.g. the first argument of `substr` is the data, but it is the third argument of `gsub`).
* It is clear in the `tidyverse` code where the tidying begins and ends.
* Each step of tidying is well defined (e.g. `gather %>% mutate %>% filter %>% select`).
* `who2` is only created once, while we have to modify `who3` several times before the data are tidy.
* String parsing is much more readable when using the pipes.

## Mutate

We saw an example of `mutate` above. In general, using the `dplyr` package will result in cleaner, simpler code, and it will save you time (e.g. because your code will be less buggy). There is no real equivalent to `mutate` in base R. 

* `mutate` allows you to add all new variables in one construct, rather than spreading them out over multiple lines of code.
* `mutate` (as well as `data_frame`) allows you to use variables you have created in the same function call (e.g. we created variables `group` and `sexAgeCat` above and were able to use them to create other variables in the same call to `mutate`).

## tibbles

In addition to what has already been discussed, `tibble` has a few more features I would like to highlight:

* Variables in `data.frame` are limited to vectors, but in `data_frame` you can have a list (e.g. `who2$group` was a list of string vectors).
* `tibble` is more computationally efficient.
* Import functions (e.g. `read_csv`) are more computationally efficient.

Click [here](blob/master/1-Tidyverse.md) for more information and examples.

## Miscelaneous data processing tools in the `tidyverse`

### lubridate

The [lubridate](https://github.com/tidyverse/lubridate/blob/master/vignettes/lubridate.Rmd) package has some additional functions that will help us work with dates and time.

- `now()` essentially does the same thing as `Sys.time()` by default, but you can optionally specify a different timezone.
- `second()`, `minute()`, `hour()`, `day()`, `year()`, ... allow you to extract or set (change) the specified part of a POSIXlt formatted variable.
- `seconds()`, `minutes()`, `hours()`, `days()`, `years()`, ... allow you to specify a period of time (e.g. `days(5)` is 5 days).

Examples:

```{r lubridate examples}
require(lubridate)

# by default, these both produce the same result
Sys.time()
now()

# however, now() can be used to get the time in a different timezone
now("GMT")

#### the Wall Street Market crash of 1929
crash <- strptime("Oct 29, 1929 9:30 AM", format = "%B %d, %Y %H:%M %p")

# what day of the week was the Wall Street Market crash of 1929?
wday(crash, label = TRUE)

# really the crash started on Monday and continued into Tuesday
# this is the interval over which the crash happened
crash <- interval(crash - days(1),                        # Monday @ 9:30 AM
                  crash + period(hours = 6, minutes = 30))# Tuesday @ 4:00 PM


#### arithmetic can be interesting
                    # leap year    not leap year
jan28 <- strptime(c("2016-01-28", "2017-01-28"), format = "%Y-%m-%d")

# we can add a month and then a day is OK
jan28 + months(1) + days(1)

# adding a day then a month can be problematic
jan28 + days(1) + months(1)
```

Note: there seems to be a bug with R's handling of the default time zone settings on some systems (as of early 2018). If you happen to run into this error,

```
Error in (function (dt, year, month, yday, mday, wday, hour, minute, second,  : 
  Invalid timezone of input vector: ""
```

this bit of code will set things straight (or something similar, e.g. EDT or GMT):

```{r eval = FALSE}
Sys.setenv(TZ="EST")
```

### stringr

The [stringr]() package reproduces many of the same functions from base R, but their names and arguments are much more consistent (e.g. the input string is always the first argument, thus they work better with `%>%`). Some examples that I like include:

* Instead of the single function, `grep`, `stringr` employs three different functions - I find this easier to read and easier to use.
    * `str_detect` is the equivalent of `grepl`.
    * `str_which` is the equivalent of the default `grep`.
    * `str_subset` is the equivalent of `grep(..., value = TRUE)`.
* `str_split_fixed` returns a matrix of substrings (`str_split` is equivalent to `strsplit`).
* `str_to_title` selectively capitalizes the first letter of each word - suitable for using as a title.

Examples:

```{r stringr examples}
require(stringr)

# group names from the who data set that we parsed above
strings <- names(who)[5:9]

# Find all columns from the 0-14 age group
grepl("014", strings)
str_detect(strings, "014")

grep("014", strings)
str_which(strings, "014")

grep("014", strings, value = TRUE)
str_subset(strings, "014")

# remove "new_" from the string and split on _ (return list)
gsub("new_?", "", strings) %>%
    strsplit("_")

str_replace(strings, "new_?", "") %>%
    str_split("_")

# remove "new_" from the string and split on _ (return matrix)
gsub("new_?", "", strings) %>%
    strsplit("_") %>%
    unlist() %>%
    matrix(ncol = 2, byrow = TRUE)    

str_replace(strings, "new_?", "") %>%
    str_split_fixed("_", 2)
```